---
title: "Density-on-scalar regression for class performence"
author: "Johannes Wagner"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: yes
    toc_depth: 4
    code_folding: hide
---
  
```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=F, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```

```{r echo=FALSE, include=FALSE}
library(tidyverse)
library(doBy)
library(broom)
library("data.table")
library(magrittr) 
library(gtools) # various programming tools
library("robCompositions") # for compData analysis
library(compositions)
library(ggplot2) # for plots
library(gridExtra) # for several plots in one window
library(expss) # to assign var labels
library(sm) # for plotting densities
```


# Aim of work 



# Dataset: STAR

## Intro and key facts

## Relevance

### Experimental design

## Research field

## Data preparation

The first step is to load the STAR dataset and recode relavent variables for the purposes of our analysis.

```{r echo=TRUE}
# rename our dataset  
load("/home/johannes/Documents/Projects/2020_Master_STAR/master20/data/STAR_Students.RData")
# rename and drop observation with only NAs
main_data <- x %>% filter(!is.na(stdntid))
```


### Exploratory variables

#### Entry Levels for each student

Krüger controls for entry levels and schools, since random assignment took place within schools and there are several arguments that entry level has an effect on the strength of the treatment variable.

```{r echo=TRUE}
main_data <- 
  main_data %>% 
  mutate(entry_level = ifelse(X=="YES",1,0)) %>% 
    mutate(entry_level = ifelse(X.1=="YES" & X=="NO",2,entry_level)) %>% 
      mutate(entry_level = ifelse(X.2=="YES" & X=="NO" & X.1=="NO",3,entry_level))  %>% 
        mutate(entry_level = ifelse(X.3=="YES" & X=="NO" & X.1=="NO" & X.2=="NO",4,entry_level)) 
main_data$entry_level <- factor(main_data$entry_level, labels=c("Kindergarten", "1st Grade", "2nd Grade", "3rd Grade"))
# create Fixed effects for school-by-entry-wave effects (see Krüger 2001)
# 4 levels

main_data <- 
  main_data %>% mutate(school_at_entry = ifelse(entry_level=="Kindergarten", gkschid, 0)) %>%
    mutate(school_at_entry = ifelse(entry_level=="1st Grade", g1schid, school_at_entry)) %>%
      mutate(school_at_entry = ifelse(entry_level=="2nd Grade", g2schid, school_at_entry)) %>%
        mutate(school_at_entry = ifelse(entry_level=="3rd Grade", g3schid, school_at_entry))
main_data$school_at_entry <- factor(main_data$school_at_entry) 
# attributes(main_data$school_at_entry)
# 79 levels
```

Create Krügers school by entry fixed effects:

```{r echo=TRUE}
# make a factor variable, which combines the two informations to get an interaction effect
# this is "school_by_entry_wave" variable (see Krüger 1999)
main_data$school_by_entry <- with(main_data, interaction(school_at_entry, entry_level))
```


#### Race

The variable `race` is coded into 3 factors: White/Asian, Black, Other

At first, for students:

```{r echo=TRUE}
# Change the levels of var race
main_data$race <-  recode_factor(main_data$race, `WHITE`="WHITE/ASIAN", `BLACK`="BLACK", `ASIAN`="WHITE/ASIAN", .default = "OTHER")
summary(main_data$race)
```

Secondly, for teachers:

```{r echo=TRUE}
# Change the levels of var race for Teacher in Kindergarten
main_data$gktrace <-  recode_factor(main_data$gktrace, `WHITE`="WHITE/ASIAN", `BLACK`="BLACK", `ASIAN`="WHITE/ASIAN", .default = "OTHER")

# and for 1st,2nd and 3rd Grade
main_data$g1trace <-  recode_factor(main_data$g1trace, `WHITE`="WHITE/ASIAN", `BLACK`="BLACK", `ASIAN`="WHITE/ASIAN", .default = "OTHER")
main_data$g2trace <-  recode_factor(main_data$g2trace, `WHITE`="WHITE/ASIAN", `BLACK`="BLACK", `ASIAN`="WHITE/ASIAN", .default = "OTHER")
main_data$g3trace <-  recode_factor(main_data$g3trace, `WHITE`="WHITE/ASIAN", `BLACK`="BLACK", `ASIAN`="WHITE/ASIAN", .default = "OTHER")

# Give statistics for Kindergarten Teachers
summary(main_data$gktrace)
```

#### Free lunch 

This recodes the grade specific free lunch variables into one Dummy indicating overall `free lunch` status.

```{r echo=TRUE}
main_data <-  main_data %>% mutate(lunch = ifelse(entry_level=="Kindergarten" & gkfreelunch=="FREE LUNCH",1, ifelse(entry_level=="1st Grade" & g1freelunch=="FREE LUNCH",1, ifelse(entry_level=="2nd Grade" & g2freelunch=="FREE LUNCH", 1,ifelse(entry_level=="3rd Grade" & g3freelunch=="FREE LUNCH",1 , 0)))))   

summary(main_data$lunch)
```

#### Classtype

The three classtypes are recoded as a dummy variable. This is just done for easier computation of later regression models and else.

We compute the actual classtype by entree wave^[This procedure is advisable since some students switch their assigned classtype during their school time, which violates the central assumption of random assignment. Therefore it makes sense to check for the entry level assigment and not the actual class. The number of those cases is very small though (see Krüger 2001)]:


```{r echo=TRUE}
main_data <- 
  main_data %>% mutate(smallDu = ifelse(entry_level=="Kindergarten" & gkclasstype=="SMALL CLASS",1 , ifelse(entry_level=="1st Grade" & g1classtype=="SMALL CLASS",1, ifelse(entry_level=="2nd Grade" & g2classtype=="SMALL CLASS",1, ifelse(entry_level=="3rd Grade" & g3classtype=="SMALL CLASS",1,0)))))

main_data <- 
  main_data %>% mutate(regAidDu = ifelse(entry_level=="Kindergarten" & gkclasstype=="REGULAR + AIDE CLASS",1 , ifelse(entry_level=="1st Grade" & g1classtype=="REGULAR + AIDE CLASS",1, ifelse(entry_level=="2nd Grade" & g2classtype=="REGULAR + AIDE CLASS",1, ifelse(entry_level=="3rd Grade" & g3classtype=="REGULAR + AIDE CLASS",1,0)))))

summary(main_data$smallDu); summary(main_data$regAidDu)
```

#### Teachers education

Following Krüger, teacher education is recoded as a DUMMY for Master degree or higher. Summary Statistics are displayed for each grade (k-3):

```{r echo=TRUE}
# create a Dummy for Master or Higher
main_data <- 
  main_data %>% 
    mutate(gktdegree = ifelse(gkthighdegree=="MASTERS" |gkthighdegree=="MASTER +" | gkthighdegree=="SPECIALIST" |gkthighdegree=="DOCTORAL",1,0), g1tdegree = ifelse(g1thighdegree=="MASTERS" |g1thighdegree=="MASTER +" | g1thighdegree=="SPECIALIST" |g1thighdegree=="DOCTORAL",1,0), g2tdegree = ifelse(g2thighdegree=="MASTERS" |g2thighdegree=="MASTER +" | g2thighdegree=="SPECIALIST" |g2thighdegree=="DOCTORAL",1,0), g3tdegree = ifelse(g3thighdegree=="MASTERS" |g3thighdegree=="MASTER +" | g3thighdegree=="SPECIALIST" |g3thighdegree=="DOCTORAL",1,0))

summary(main_data$gktdegree); summary(main_data$g1tdegree); summary(main_data$g2tdegree); summary(main_data$g3tdegree)
```



#### School ID

School ID is recoded as a factor variable.

```{r echo=TRUE}
main_data$gkschid <- factor(main_data$gkschid)
main_data$g1schid <- factor(main_data$g1schid)
main_data$g2schid <- factor(main_data$g2schid)
main_data$g3schid <- factor(main_data$g3schid)
# attributes(main_data$gkschid)

attributes(main_data$gkschid)
```


### Dependent variable

Test scores are usually the measure of choice for evaluating school performance of students. As already said, the aim of the work is to identify factors that influence school performance. In contrast to previous studies the entire distribution of test performance within each class is to be considered. To quantify the relative test performance, it is advisable to use percentiles (@Krüger 2001).

As a first step, the average values of the test results for the subjects math, reading and writing are calculated. This can later be used as a raw (and possible problematic) measurement of a students ability to perform in tests.

```{r echo=TRUE}
# Compute Raw average Scores as a new Variable of Interest
main_data <- 
  main_data %>% mutate(gkaversc = (gktmathss + gktreadss + gkwordskillss)/3, g1aversc = (g1tmathss + g1treadss + g1wordskillss)/3, g2aversc = (g2tmathss + g2treadss + g2wordskillss)/3, g3aversc = (g3tmathss + g3treadss + g3wordskillss)/3)  
```

For easier data manipulation the dataset is split by student grades, which can either be kindergarden (*gk*), first grade (*g1*), second grade (*g2*) or third grade (*g3*).  

```{r echo=TRUE}
main_data_gk <- 
  main_data %>% filter(X=="YES")
main_data_g1 <- 
  main_data %>% filter(X=="YES" | X.1=="YES")
main_data_g2 <- 
  main_data %>% filter(X=="YES" | X.1=="YES" | X.2=="YES")
main_data_g3 <- 
  main_data %>% filter(X=="YES" | X.1=="YES" | X.2=="YES" | X.3=="YES")
```  

#### Percentile ranks

In the following, the calculation steps for calculating ranking positions are replicated according to the Krüger template. Brief explanations can be found inside the code.

Percentage ranks are first computed for all students in regular classes and then the students, who were assigned to small classes, are given their ranks **relative** to the ranks of the distribution of regular students. Test scores are available for math, reading and word comprehension. Following, the mean values for each group of students and each subject are displayed. The `percent_rank()` function from the `dplyr` package was used for the computation of `perc_[subject]_[grade]` variables. The function gives the lower percentile rank to students with equal score.^[This explains why the mean values for regular students (baseline) are lower than 0.5.]  

In this version, I computed percentiles for average test results differently from Krügers suggestion. Instead of calculating the rank of a students as the average of his or her subject specific ranks, the average of the subject-specific test scores was computed and then each (regular) got assigned to his or her school specific rank. 

##### Percentile rank (Krüger)

Due to the nature of percentiles the rank distribution of regular students should follow a uniform distribution. 

The mean scores conditional on classtype and grade are displayed (cp. Krüger (1999), p. 508)). The first moment of the score distribution supports the idea, that students in small classes achieve higher test results than students in regular classes. 

```{r echo=TRUE}
# See Krüger_199, p. 506ff
# select data from students in regular classes without any missing scores
reg_scores_gk <- 
  main_data_gk %>% filter(gkclasstype=="REGULAR CLASS" | gkclasstype=="REGULAR + AIDE CLASS", !is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) ) %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc)

# Select data from students in small classes
small_scores_gk <- 
  main_data_gk %>% filter(gkclasstype=="SMALL CLASS", !is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) )  %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc  )

# calculate the percentile ranks for each subject and for the average score from the raw average test result
# This is done for every student in a regular class
reg_scores_gk <- 
  reg_scores_gk %>% mutate(perc_maths_gk=percent_rank(gktmathss), perc_read_gk =percent_rank(gktreadss), perc_word_gk=percent_rank(gkwordskillss), perc_av_gk=percent_rank(gkaversc))

# Following Krüger, we calculate the percentile scores for students in small classes relative to the population of students in regular classes
scores_gk <- reg_scores_gk

for (i in 1:nrow(small_scores_gk)) {
  # select ith observation
  obs <- small_scores_gk[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_gk, obs) %>% 
    mutate(perc_maths_gk=percent_rank(gktmathss), perc_read_gk =percent_rank(gktreadss), perc_word_gk=percent_rank(gkwordskillss), perc_av_gk=percent_rank(gkaversc))
  # add observation to merged scores
  scores_gk <-  add_row(scores_gk, tail(temp,1))
}


# Calculate Summary Statistics (See: Krüger (1999), p. 508)
scores_gk %>% 
  filter(gkclasstype=="REGULAR CLASS" | gkclasstype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_gk), mean_math = mean(perc_maths_gk), mean_read = mean(perc_read_gk), mean_word = mean(perc_word_gk))
```

Mean Scores for students in small classes in Kindergarten: 

```{r echo=TRUE}
scores_gk %>% 
  filter(gkclasstype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_gk), mean_math = mean(perc_maths_gk), mean_read = mean(perc_read_gk), mean_word = mean(perc_word_gk))    
```

The densities for both subgroups are shown below:

```{r echo=TRUE}
# make a factor variable for small and regular
scores_gk <- scores_gk %>%
  mutate(classtype = ifelse(gkclasstype=="SMALL CLASS",1,0))
class_factor  <- factor(scores_gk$classtype, levels = c(0,1), labels = c("0 Regular", "1 SMALL")) 


# plot densities
sm.density.compare(scores_gk$perc_maths_gk, scores_gk$classtype, xlab="Student percentile rank for math")
title(main="Percentile distribution by classtype")
# add legend
colfill<-c(2:(2+length(levels(class_factor))))
legend("bottomleft", levels(class_factor), fill=colfill) 
```

Obviously this plot is not optimal but supports the idea that a small class assignment shifts the distribution to the right. 
We do see that the distribution of regular students does not follow an exact uniform distribution, even if we would constrain the values on the 0-1 interval. That is due to the fact that `percent_rank` uses the *ties method* "min" where equal scores are assigned to the minimal percentile.^[Four example, if we would have five students with student 1 having the lowest score, student 2 the second lowest score and so on, they would be assigned to the percentiles: 0,0.25,0.5,0.75,1. But if student 4 and 5 have the same scores then they get assigned into the same percentile. In this case this would be the 0.75 percentile and the percentile ranks for the five students would be 0,0.25,0.5,0.75,0.75. This results in little peaks where many students have the same scores and those peaks are somewhat shifted to the left compared to their theoretical position in the underlying uniform distribution.] 


The same computational steps are repeated for 1st, 2nd and 3rd Grade and are saved in datasets `score_[grade]`

Mean Scores for students in regular classes in Grade 1: 

```{r echo=TRUE}
reg_scores_g1 <- 
  main_data_g1 %>% filter(g1classtype=="REGULAR CLASS" | g1classtype=="REGULAR + AIDE CLASS", !is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc)

small_scores_g1 <- 
  main_data_g1 %>% filter(g1classtype=="SMALL CLASS", !is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc)

reg_scores_g1 <- 
  reg_scores_g1 %>% mutate(perc_maths_g1=percent_rank(g1tmathss), perc_read_g1 =percent_rank(g1treadss), perc_word_g1=percent_rank(g1wordskillss), perc_av_g1=percent_rank(g1aversc))

scores_g1 <- reg_scores_g1

for (i in 1:nrow(small_scores_g1)) {
  # select ith observation
  obs <- small_scores_g1[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g1, obs) %>% 
    mutate(perc_maths_g1=percent_rank(g1tmathss), perc_read_g1 =percent_rank(g1treadss), perc_word_g1=percent_rank(g1wordskillss), perc_av_g1=percent_rank(g1aversc))
  # add observation to merged scores
  scores_g1 <-  add_row(scores_g1, tail(temp,1))
}

scores_g1 %>% 
  filter(g1classtype=="REGULAR CLASS" | g1classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_g1), mean_math = mean(perc_maths_g1), mean_read = mean(perc_read_g1), mean_word = mean(perc_word_g1))

```

Mean Scores for students in small classes in Grade 1: 

```{r echo=TRUE}
scores_g1 %>% 
  filter(g1classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_g1), mean_math = mean(perc_maths_g1), mean_read = mean(perc_read_g1), mean_word = mean(perc_word_g1))    
```

Mean Scores for students in regular classes in Grade 2: 

```{r echo=TRUE}
reg_scores_g2 <- 
  main_data_g2 %>% filter(g2classtype=="REGULAR CLASS" | g2classtype=="REGULAR + AIDE CLASS", !is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc )

small_scores_g2 <- 
  main_data_g2 %>% filter(g2classtype=="SMALL CLASS", !is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc)

reg_scores_g2 <- 
  reg_scores_g2 %>% mutate(perc_maths_g2=percent_rank(g2tmathss), perc_read_g2 =percent_rank(g2treadss), perc_word_g2=percent_rank(g2wordskillss), perc_av_g2=percent_rank(g2aversc))

scores_g2 <- reg_scores_g2

for (i in 1:nrow(small_scores_g2)) {
  # select ith observation
  obs <- small_scores_g2[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g2, obs) %>% 
    mutate(perc_maths_g2=percent_rank(g2tmathss), perc_read_g2 =percent_rank(g2treadss), perc_word_g2=percent_rank(g2wordskillss),
perc_av_g2=percent_rank(g2aversc))
  # add observation to merged scores
  scores_g2 <-  add_row(scores_g2, tail(temp,1))
}


scores_g2 %>% 
  filter(g2classtype=="REGULAR CLASS" | g2classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_g2), mean_math = mean(perc_maths_g2), mean_read = mean(perc_read_g2), mean_word = mean(perc_word_g2))

```

Mean Scores for students in small classes in Grade 2:

```{r echo=TRUE}
scores_g2 %>% 
  filter(g2classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_g2, na.rm = TRUE), mean_math = mean(perc_maths_g2), mean_read = mean(perc_read_g2), mean_word = mean(perc_word_g2))
```

Mean Scores for students in regular classes in Grade 3:


```{r echo=TRUE}
reg_scores_g3 <- 
  main_data_g3 %>% filter(g3classtype=="REGULAR CLASS" | g3classtype=="REGULAR + AIDE CLASS", !is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid,g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid,g3aversc  )

small_scores_g3 <- 
  main_data_g3 %>% filter(g3classtype=="SMALL CLASS", !is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid, g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid,g3aversc  )

reg_scores_g3 <- 
  reg_scores_g3 %>% mutate(perc_maths_g3=percent_rank(g3tmathss), perc_read_g3 =percent_rank(g3treadss), perc_word_g3=percent_rank(g3wordskillss),  perc_av_g3=percent_rank(g3aversc))

scores_g3 <- reg_scores_g3

for (i in 1:nrow(small_scores_g3)) {
  # select ith observation
  obs <- small_scores_g3[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g3, obs) %>% 
    mutate(perc_maths_g3=percent_rank(g3tmathss), perc_read_g3 =percent_rank(g3treadss), perc_word_g3=percent_rank(g3wordskillss),  perc_av_g3=percent_rank(g3aversc))
  # add observation to merged scores
  scores_g3 <-  add_row(scores_g3, tail(temp,1))
}

scores_g3 %>% 
  filter(g3classtype=="REGULAR CLASS" | g3classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_g3), mean_math = mean(perc_maths_g3), mean_read = mean(perc_read_g3), mean_word = mean(perc_word_g3))

```
Mean Scores for students in small classes in Grade 3:

```{r echo=TRUE}
scores_g3 %>% 
  filter(g3classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_g3), mean_math = mean(perc_maths_g3), mean_read = mean(perc_read_g3), mean_word = mean(perc_word_g3))    
```

Finally, all percentile ranks are merged by student ID to get a dataset, which contails the ranks of every student in each grade:

```{r echo=TRUE}
# join all scores
scores_all <- 
  scores_gk %>% 
    full_join(scores_g1, by = "stdntid") %>% 
      full_join(scores_g2, by = "stdntid") %>% 
        full_join(scores_g3, by = "stdntid") %>% 
          select(stdntid, perc_av_g3, perc_maths_g3, perc_read_g3, perc_word_g3, perc_av_g2, perc_maths_g2, perc_read_g2, perc_word_g2, perc_av_g1, perc_maths_g1, perc_read_g1, perc_word_g1,perc_av_gk, perc_maths_gk, perc_read_gk, perc_word_gk)
# merge them with the main dataset
main_data <- 
  main_data %>% full_join(scores_all, by="stdntid")
```



#### Population percentile ranks

This approach computes the variables `perc2_[subject]_[grade]` directly from the whole population. 

Mean Scores for students in small classes in Kindergarden:

```{r echo=TRUE}
# Kindergarden
scores2_gk <- 
  main_data_gk %>% 
    filter(!is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) )  %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc) %>%   
      mutate(perc_fp_maths_gk=percent_rank(gktmathss), perc_fp_read_gk =percent_rank(gktreadss), perc_fp_word_gk=percent_rank(gkwordskillss), perc_fp_av_gk=percent_rank(gkaversc))

scores2_gk %>% 
  filter(gkclasstype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_gk), mean_math = mean(perc_fp_maths_gk), mean_read = mean(perc_fp_read_gk), mean_word = mean(perc_fp_word_gk))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_gk %>% 
  filter(gkclasstype=="REGULAR CLASS" | gkclasstype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_gk), mean_math = mean(perc_fp_maths_gk), mean_read = mean(perc_fp_read_gk), mean_word = mean(perc_fp_word_gk))
```  

Mean Scores for students in small classes in Grade 1:

```{r echo=TRUE}
# Kindergarden
scores2_g1 <- 
  main_data_g1 %>% 
    filter(!is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc) %>%   
      mutate(perc_fp_maths_g1=percent_rank(g1tmathss), perc_fp_read_g1 =percent_rank(g1treadss), perc_fp_word_g1=percent_rank(g1wordskillss), perc_fp_av_g1=percent_rank(g1aversc))

scores2_g1 %>% 
  filter(g1classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g1), mean_math = mean(perc_fp_maths_g1), mean_read = mean(perc_fp_read_g1), mean_word = mean(perc_fp_word_g1))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_g1 %>% 
  filter(g1classtype=="REGULAR CLASS" | g1classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g1), mean_math = mean(perc_fp_maths_g1), mean_read = mean(perc_fp_read_g1), mean_word = mean(perc_fp_word_g1))
``` 

Mean Scores for students in small classes in Grade 2:

```{r echo=TRUE}
# Kindergarden
scores2_g2 <- 
  main_data_g2 %>% 
    filter(!is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc) %>%   
      mutate(perc_fp_maths_g2=percent_rank(g2tmathss), perc_fp_read_g2 =percent_rank(g2treadss), perc_fp_word_g2=percent_rank(g2wordskillss), perc_fp_av_g2=percent_rank(g2aversc))

scores2_g2 %>% 
  filter(g2classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g2), mean_math = mean(perc_fp_maths_g2), mean_read = mean(perc_fp_read_g2), mean_word = mean(perc_fp_word_g2))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_g2 %>% 
  filter(g2classtype=="REGULAR CLASS" | g2classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g2), mean_math = mean(perc_fp_maths_g2), mean_read = mean(perc_fp_read_g2), mean_word = mean(perc_fp_word_g2))
```  

Mean Scores for students in small classes in grade 3:

```{r echo=TRUE}
# Kindergarden
scores2_g3 <- 
  main_data_g3 %>% 
    filter(!is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid,g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid, g3aversc) %>%   
      mutate(perc_fp_maths_g3=percent_rank(g3tmathss), perc_fp_read_g3 =percent_rank(g3treadss), perc_fp_word_g3=percent_rank(g3wordskillss), perc_fp_av_g3=percent_rank(g3aversc))

scores2_g3 %>% 
  filter(g3classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g3), mean_math = mean(perc_fp_maths_g3), mean_read = mean(perc_fp_read_g3), mean_word = mean(perc_fp_word_g3))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_g3 %>% 
  filter(g3classtype=="REGULAR CLASS" | g3classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g3), mean_math = mean(perc_fp_maths_g3), mean_read = mean(perc_fp_read_g3), mean_word = mean(perc_fp_word_g3))
```  

Again, all percentile ranks are merged by student ID to get a dataset, which contains the ranks of every student in each grade:

```{r echo=TRUE}
# join all scores
scores_all2 <- 
  scores2_gk %>% 
    full_join(scores2_g1, by = "stdntid") %>% 
      full_join(scores2_g2, by = "stdntid") %>% 
        full_join(scores2_g3, by = "stdntid") %>% 
          select(stdntid, perc_fp_av_g3, perc_fp_maths_g3, perc_fp_read_g3, perc_fp_word_g3, perc_fp_av_g2, perc_fp_maths_g2, perc_fp_read_g2, perc_fp_word_g2, perc_fp_av_g1, perc_fp_maths_g1, perc_fp_read_g1, perc_fp_word_g1,perc_fp_av_gk, perc_fp_maths_gk, perc_fp_read_gk, perc_fp_word_gk)
# merge them with the main dataset
main_data <- 
  main_data %>% full_join(scores_all2, by="stdntid")
```


#### Percentile Ranks per School

This approach diverges from Krüger et al. 2001, but might be helpful since it directly controls variation on the school level. 

We expect percentile ranks for regular students within schools to follow a uniform distribution. Again the Variation in distribution of test scores from students in small classes is of interest.

The computation is the same as before, but this time the data is grouped by school ID.

```{r echo=TRUE}
# select relevant variables
small_scores_gk_sc <- 
  main_data_gk %>% filter(gkclasstype=="SMALL CLASS", !is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) )  %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc  )

# We simply repeat the procedure with data grouped by school 
reg_scores_gk_sc <- 
  reg_scores_gk %>%
    group_by(gkschid) %>% 
      mutate(perc_maths_gk_sc=percent_rank(gktmathss), perc_read_gk_sc =percent_rank(gktreadss), perc_word_gk_sc=percent_rank(gkwordskillss), perc_av_gk_sc=percent_rank(gkaversc)) %>%
        ungroup()

scores_gk_sc <- reg_scores_gk_sc

for (i in 1:nrow(small_scores_gk_sc)) {
  # select ith observation
  obs <- small_scores_gk_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_gk_sc, obs) %>%
    group_by(gkschid) %>%
    mutate(perc_maths_gk_sc=percent_rank(gktmathss), perc_read_gk_sc =percent_rank(gktreadss), perc_word_gk_sc=percent_rank(gkwordskillss), perc_av_gk_sc=percent_rank(gkaversc)) %>% ungroup()
  # add observation to merged scores
  scores_gk_sc <-  add_row(scores_gk_sc, tail(temp,1))
}

# Grade 1
small_scores_g1_sc <- 
  main_data_g1 %>% filter(g1classtype=="SMALL CLASS", !is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc)

reg_scores_g1_sc <- 
  reg_scores_g1 %>%
    group_by(g1schid) %>% 
      mutate(perc_maths_g1_sc=percent_rank(g1tmathss), perc_read_g1_sc =percent_rank(g1treadss), perc_word_g1_sc=percent_rank(g1wordskillss), perc_av_g1_sc=percent_rank(g1aversc))  %>%
        ungroup()

scores_g1_sc <- reg_scores_g1_sc

for (i in 1:nrow(small_scores_g1_sc)) {
  # select ith observation
  obs <- small_scores_g1_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g1_sc, obs) %>%
    group_by(g1schid) %>%
    mutate(perc_maths_g1_sc=percent_rank(g1tmathss), perc_read_g1_sc =percent_rank(g1treadss), perc_word_g1_sc=percent_rank(g1wordskillss), perc_av_g1_sc=percent_rank(g1aversc))  %>%
        ungroup()
  # add observation to merged scores
  scores_g1_sc <-  add_row(scores_g1_sc, tail(temp,1))
}

# Grade 2
small_scores_g2_sc <- 
  main_data_g2 %>% filter(g2classtype=="SMALL CLASS", !is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc  )

reg_scores_g2_sc <- 
  reg_scores_g2 %>%
    group_by(g2schid) %>% 
      mutate(perc_maths_g2_sc=percent_rank(g2tmathss), perc_read_g2_sc =percent_rank(g2treadss), perc_word_g2_sc=percent_rank(g2wordskillss), perc_av_g2_sc=percent_rank(g2aversc))  %>%
        ungroup()

scores_g2_sc <- reg_scores_g2_sc

for (i in 1:nrow(small_scores_g2_sc)) {
  # select ith observation
  obs <- small_scores_g2_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g2_sc, obs) %>%
    group_by(g2schid) %>%
    mutate(perc_maths_g2_sc=percent_rank(g2tmathss), perc_read_g2_sc =percent_rank(g2treadss), perc_word_g2_sc=percent_rank(g2wordskillss), perc_av_g2_sc=percent_rank(g2aversc))  %>%
        ungroup()
  # add observation to merged scores
  scores_g2_sc <-  add_row(scores_g2_sc, tail(temp,1))
}


# Grade 3
small_scores_g3_sc <- 
  main_data_g3 %>% filter(g3classtype=="SMALL CLASS", !is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid,g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid , g3aversc )

reg_scores_g3_sc <- 
  reg_scores_g3 %>%
    group_by(g3schid) %>% 
      mutate(perc_maths_g3_sc=percent_rank(g3tmathss), perc_read_g3_sc =percent_rank(g3treadss), perc_word_g3_sc=percent_rank(g3wordskillss), perc_av_g3_sc=percent_rank(g3aversc))  %>%
        ungroup()

scores_g3_sc <- reg_scores_g3_sc

for (i in 1:nrow(small_scores_g3_sc)) {
  # select ith observation
  obs <- small_scores_g3_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g3_sc, obs) %>%
    group_by(g3schid) %>%
    mutate(perc_maths_g3_sc=percent_rank(g3tmathss), perc_read_g3_sc =percent_rank(g3treadss), perc_word_g3_sc=percent_rank(g3wordskillss), perc_av_g3_sc=percent_rank(g3aversc))  %>%
        ungroup()
  # add observation to merged scores
  scores_g3_sc <-  add_row(scores_g3_sc, tail(temp,1))
}


scores_all_sc <- 
  scores_gk_sc %>% 
    full_join(scores_g1_sc, by = "stdntid") %>% 
      full_join(scores_g2_sc, by = "stdntid") %>% 
        full_join(scores_g3_sc, by = "stdntid") %>% 
          select(stdntid, perc_av_g3_sc, perc_maths_g3_sc, perc_read_g3_sc, perc_word_g3_sc, perc_av_g2_sc, perc_maths_g2_sc, perc_read_g2_sc, perc_word_g2_sc, perc_av_g1_sc, perc_maths_g1_sc, perc_read_g1_sc, perc_word_g1_sc,perc_av_gk_sc, perc_maths_gk_sc, perc_read_gk_sc, perc_word_gk_sc)
# we get a problem here, because covariates are in different categories, better make a dataset with new variables and merge them with the main_data
main_data <- 
  main_data %>% full_join(scores_all_sc, by="stdntid")
```

### Saving the dataset

The the code below variable labels are assigned to the new variables.

```{r echo=TRUE}
# Average test scores
attributes(main_data$gkaversc) <- list(comment = "Average test score in kindergarten")   
attributes(main_data$g1aversc) <- list(comment = "Average test score in Grade1") 
attributes(main_data$g2aversc) <- list(comment = "Average test score in Grade2") 
attributes(main_data$g3aversc) <- list(comment = "Average test score in Grade3") 
# percentiles of whole distribution
attributes(main_data$perc_maths_gk) <- list(comment = "Percentile Math over all STAR students in kindergarten")   
attributes(main_data$perc_maths_g1) <- list(comment = "Percentile Math over all STAR students in Grade1")   
attributes(main_data$perc_maths_g2) <- list(comment = "Percentile Math over all STAR students in Grade2")   
attributes(main_data$perc_maths_g3) <- list(comment = "Percentile Math over all STAR students in Grade3")   
attributes(main_data$perc_read_gk) <- list(comment = "Percentile Reading over all STAR students in kindergarten")   
attributes(main_data$perc_read_g1) <- list(comment = "Percentile Reading over all STAR students in Grade1")   
attributes(main_data$perc_read_g2) <- list(comment = "Percentile Reading over all STAR students in Grade2")   
attributes(main_data$perc_read_g3) <- list(comment = "Percentile Reading over all STAR students in Grade3")  
attributes(main_data$perc_word_gk) <- list(comment = "Percentile Word comprehension over all STAR students in kindergarten")   
attributes(main_data$perc_word_g1) <- list(comment = "Percentile Word comprehension over all STAR students in Grade1")   
attributes(main_data$perc_word_g2) <- list(comment = "Percentile Word comprehension over all STAR students in Grade2")   
attributes(main_data$perc_word_g3) <- list(comment = "Percentile Word comprehension over all STAR students in Grade3") 
attributes(main_data$perc_av_gk) <- list(comment = "Percentile average score over all STAR students in kindergarten")   
attributes(main_data$perc_av_g1) <- list(comment = "Percentile average score  over all STAR students in Grade1")   
attributes(main_data$perc_av_g2) <- list(comment = "Percentile average score  over all STAR students in Grade2")   
attributes(main_data$perc_av_g3) <- list(comment = "Percentile average score  over all STAR students in Grade3") 
# percentiles within schools
attributes(main_data$perc_maths_gk_sc) <- list(comment = "Percentile Math within schools in kindergarten")   
attributes(main_data$perc_maths_g1_sc) <- list(comment = "Percentile Math within schools in Grade1")   
attributes(main_data$perc_maths_g2_sc) <- list(comment = "Percentile Math within schools in Grade2")   
attributes(main_data$perc_maths_g3_sc) <- list(comment = "Percentile Math within schools in Grade3")   
attributes(main_data$perc_read_gk_sc) <- list(comment = "Percentile Reading within schools in kindergarten")   
attributes(main_data$perc_read_g1_sc) <- list(comment = "Percentile Reading within schools in Grade1")   
attributes(main_data$perc_read_g2_sc) <- list(comment = "Percentile Reading within schools in Grade2")   
attributes(main_data$perc_read_g3_sc) <- list(comment = "Percentile Reading within schools in Grade3")  
attributes(main_data$perc_word_gk_sc) <- list(comment = "Percentile Word comprehension within schools in kindergarten")   
attributes(main_data$perc_word_g1_sc) <- list(comment = "Percentile Word comprehension within schools in Grade1")   
attributes(main_data$perc_word_g2_sc) <- list(comment = "Percentile Word comprehension within schools in Grade2")   
attributes(main_data$perc_word_g3_sc) <- list(comment = "Percentile Word comprehension within schools in Grade3") 
attributes(main_data$perc_av_gk_sc) <- list(comment = "Percentile average score within schools in kindergarten")   
attributes(main_data$perc_av_g1_sc) <- list(comment = "Percentile average score  within schools in Grade1")   
attributes(main_data$perc_av_g2_sc) <- list(comment = "Percentile average score  within schools in Grade2")   
attributes(main_data$perc_av_g3_sc) <- list(comment = "Percentile average score  within schools in Grade3") 
# label race variable for students
attributes(main_data$race) <- list(levels = c("WHITE/ASIAN", "BLACK", "OTHER"), comment = "Student race in 3 categories", class = c("ordered", "factor"))
# label race variable for teachers by class
attributes(main_data$gktrace) <- list(levels = c("WHITE/ASIAN", "BLACK", "OTHER"), comment = "Teacher race kindergarten in 3 categories", class = c("ordered", "factor"))
attributes(main_data$g1trace) <- list(levels = c("WHITE/ASIAN", "BLACK", "OTHER"), comment = "Teacher race grade1 in 3 categories", class = c("ordered", "factor"))
attributes(main_data$g2trace) <- list(levels = c("WHITE/ASIAN", "BLACK", "OTHER"), comment = "Teacher race grade2 in 3 categories", class = c("ordered", "factor"))
attributes(main_data$g3trace) <- list(levels = c("WHITE/ASIAN", "BLACK", "OTHER"), comment = "Teacher race grade3 in 3 categories", class = c("ordered", "factor"))
# label entry level
attributes(main_data$entry_level) <- list(comment = "Indikator at which grade student entered STAR", levels=levels(main_data$entry_level), class=class(main_data$entry_level))
# label school at entryy
attributes(main_data$school_at_entry) <- list(comment = "Indikator which school student was attending when entering STAR", levels=levels(main_data$school_at_entry), class=class(main_data$school_at_entry))
# label school by entree
attributes(main_data$school_by_entry) <- list(comment = "Indikator for school and entry level student when entering STAR", levels=levels(main_data$school_by_entry), class=class(main_data$school_by_entry))
# label class type dummies
attributes(main_data$smallDu) <- list(comment = "Dummy for students who entered STAR in a small class")
attributes(main_data$regAidDu) <- list(comment = "Dummy for students who entered STAR in a regular class with Aides")
# label teacher education dummy_ Master or higehr for each grade
attributes(main_data$gktdegree) <-  list(comment = "Dummy for teacher with Master degree or higher in kindergarten")
attributes(main_data$g1tdegree) <-  list(comment = "Dummy for teacher with Master degree or higher in grade1")
attributes(main_data$g2tdegree) <-  list(comment = "Dummy for teacher with Master degree or higher in grade2")
attributes(main_data$g3tdegree) <-  list(comment = "Dummy for teacher with Master degree or higher in grade3")
# label lunch dummy
attributes(main_data$lunch) <-  list(comment = "Dummy for students with FREE LUNCH status when the joined STAR")
#
```

Let's save the manipulated data sets for later access.

```{r echo=TRUE}
save(main_data, file = "/home/johannes/Documents/Projects/2020_Master_STAR/master20/data/main_data.RData")    
```


# Repetition of previous analyzes



# Conceptual framework: FAMM

# Density analysis

# Implementation with refund

# To Do's and questions