---
title: "Density-on-scalar regression for class performence"
author: "Johannes Wagner"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: yes
    toc_depth: 4
    code_folding: hide
---
  
```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=F, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```

```{r echo=FALSE, include=FALSE}
library(tidyverse)
library(doBy)
library(broom)
library("data.table")
library(magrittr) 
library(gtools) # various programming tools
library("robCompositions") # for compData analysis
library(compositions)
library(ggplot2) # for plots
library(gridExtra) # for several plots in one window
library(expss) # to assign var labels
library(sm) # for plotting densities
```


# Aim of work 



# Dataset: STAR

## Intro and key facts

## Relevance

### Experimental design

## Research field

## Data preparation

### Exploratory variables

### Dependent variable

Test scores are usually the measure of choice for evaluating school performance of students. As already said, the aim of the work is to identify factors that influence school performance. In contrast to previous studies the entire distribution of test performance within each class is to be considered. To quantify the relative test performance, it is advisable to use percentiles (@Krüger 2001).

As a first step, the average values of the test results for the subjects math, reading and writing are calculated. This can later be used as a raw (and possible problematic) measurement of a students ability to perform in tests.

```{r echo=TRUE}
# rename our dataset  
load("/home/johannes/Documents/Projects/2020_Master_STAR/master20/data/STAR_Students.RData")
# rename and drop observation with only NAs
main_data <- x %>% filter(!is.na(stdntid))
# Compute Raw average Scores as a new Variable of Interest
main_data <- 
  main_data %>% mutate(gkaversc = (gktmathss + gktreadss + gkwordskillss)/3, g1aversc = (g1tmathss + g1treadss + g1wordskillss)/3, g2aversc = (g2tmathss + g2treadss + g2wordskillss)/3, g3aversc = (g3tmathss + g3treadss + g3wordskillss)/3)  
```

For easier data manipulation the dataset is split by student grades, which can either be kindergarden (*gk*), first grade (*g1*), second grade (*g2*) or third grade (*g3*).  

```{r echo=TRUE}
main_data_gk <- 
  main_data %>% filter(X=="YES")
main_data_g1 <- 
  main_data %>% filter(X=="YES" | X.1=="YES")
main_data_g2 <- 
  main_data %>% filter(X=="YES" | X.1=="YES" | X.2=="YES")
main_data_g3 <- 
  main_data %>% filter(X=="YES" | X.1=="YES" | X.2=="YES" | X.3=="YES")
```  

#### Percentile ranks (see Krüger 2011)

In the following, the calculation steps for calculating ranking positions are replicated according to the Krüger template. Brief explanations can be found inside the code.

Percentage ranks are first computed for all students in regular classes and then the students, who were assigned to small classes, are given their ranks **relative** to the ranks of the distribution of regular students. Test scores are available for math, reading and word comprehension. Following, the mean values for each group of students and each subject are displayed. The `percent_rank()` function from the `dplyr` package was used for the computation of `perc_[subject]_[grade]` variables. The function gives the lower percentile rank to students with equal score.^[This explains why the mean values for regular students (baseline) are lower than 0.5.]  

##### Percentile rank distributions

Due to the nature of percentiles the rank distribution of regular students should follow a uniform distribution. 

The mean scores conditional on classtype and grade are displayed (cp. Krüger (1999), p. 508)). The first moment of the score distribution supports the idea, that students in small classes achieve higher test results than students in regular classes. 

```{r echo=TRUE}
# See Krüger_199, p. 506ff
# select data from students in regular classes without any missing scores
reg_scores_gk <- 
  main_data_gk %>% filter(gkclasstype=="REGULAR CLASS" | gkclasstype=="REGULAR + AIDE CLASS", !is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) ) %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc)

# Select data from students in small classes
small_scores_gk <- 
  main_data_gk %>% filter(gkclasstype=="SMALL CLASS", !is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) )  %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc  )

# calculate the percentile ranks for each subject and for the average score from the raw average test result
# This is done for every student in a regular class
reg_scores_gk <- 
  reg_scores_gk %>% mutate(perc_maths_gk=percent_rank(gktmathss), perc_read_gk =percent_rank(gktreadss), perc_word_gk=percent_rank(gkwordskillss), perc_av_gk=percent_rank(gkaversc))

# Following Krüger, we calculate the percentile scores for students in small classes relative to the population of students in regular classes
scores_gk <- reg_scores_gk

for (i in 1:nrow(small_scores_gk)) {
  # select ith observation
  obs <- small_scores_gk[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_gk, obs) %>% 
    mutate(perc_maths_gk=percent_rank(gktmathss), perc_read_gk =percent_rank(gktreadss), perc_word_gk=percent_rank(gkwordskillss), perc_av_gk=percent_rank(gkaversc))
  # add observation to merged scores
  scores_gk <-  add_row(scores_gk, tail(temp,1))
}


# Calculate Summary Statistics (See: Krüger (1999), p. 508)
scores_gk %>% 
  filter(gkclasstype=="REGULAR CLASS" | gkclasstype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_gk), mean_math = mean(perc_maths_gk), mean_read = mean(perc_read_gk), mean_word = mean(perc_word_gk))
```

Mean Scores for students in small classes in Kindergarten: 

```{r echo=TRUE}
scores_gk %>% 
  filter(gkclasstype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_gk), mean_math = mean(perc_maths_gk), mean_read = mean(perc_read_gk), mean_word = mean(perc_word_gk))    
```

The densities for both subgroups are shown below:

```{r echo=TRUE}
# make a factor variable for small and regular
scores_gk <- scores_gk %>%
  mutate(classtype = ifelse(gkclasstype=="SMALL CLASS",1,0))
class_factor  <- factor(scores_gk$classtype, levels = c(0,1), labels = c("0 Regular", "1 SMALL")) 


# plot densities
sm.density.compare(scores_gk$perc_maths_gk, scores_gk$classtype, xlab="Student percentile rank for math")
title(main="Percentile distribution by classtype")
# add legend
colfill<-c(2:(2+length(levels(class_factor))))
legend("bottomleft", levels(class_factor), fill=colfill) 
```

Obviously this plot is not optimal but supports the idea that a small class assignment shifts the distribution to the right. 
We do see that the distribution of regular students does not follow an exact uniform distribution, even if we would constrain the values on the 0-1 interval. That is due to the fact that `percent_rank` uses the *ties method* "min" where equal scores are assigned to the minimal percentile.^[Four example, if we would have five students with student 1 having the lowest score, student 2 the second lowest score and so on, they would be assigned to the percentiles: 0,0.25,0.5,0.75,1. But if student 4 and 5 have the same scores then they get assigned into the same percentile. In this case this would be the 0.75 percentile and the percentile ranks for the five students would be 0,0.25,0.5,0.75,0.75. This results in little peaks where many students have the same scores and those peaks are somewhat shifted to the left compared to their theoretical position in the underlying uniform distribution.] 


The same computational steps are repeated for 1st, 2nd and 3rd Grade and are saved in datasets `score_[grade]`

Mean Scores for students in regular classes in Grade 1: 

```{r echo=TRUE}
reg_scores_g1 <- 
  main_data_g1 %>% filter(g1classtype=="REGULAR CLASS" | g1classtype=="REGULAR + AIDE CLASS", !is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc)

small_scores_g1 <- 
  main_data_g1 %>% filter(g1classtype=="SMALL CLASS", !is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc)

reg_scores_g1 <- 
  reg_scores_g1 %>% mutate(perc_maths_g1=percent_rank(g1tmathss), perc_read_g1 =percent_rank(g1treadss), perc_word_g1=percent_rank(g1wordskillss), perc_av_g1=percent_rank(g1aversc))

scores_g1 <- reg_scores_g1

for (i in 1:nrow(small_scores_g1)) {
  # select ith observation
  obs <- small_scores_g1[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g1, obs) %>% 
    mutate(perc_maths_g1=percent_rank(g1tmathss), perc_read_g1 =percent_rank(g1treadss), perc_word_g1=percent_rank(g1wordskillss), perc_av_g1=percent_rank(g1aversc))
  # add observation to merged scores
  scores_g1 <-  add_row(scores_g1, tail(temp,1))
}

scores_g1 %>% 
  filter(g1classtype=="REGULAR CLASS" | g1classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_g1), mean_math = mean(perc_maths_g1), mean_read = mean(perc_read_g1), mean_word = mean(perc_word_g1))

```

Mean Scores for students in small classes in Grade 1: 

```{r echo=TRUE}
scores_g1 %>% 
  filter(g1classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_g1), mean_math = mean(perc_maths_g1), mean_read = mean(perc_read_g1), mean_word = mean(perc_word_g1))    
```

Mean Scores for students in regular classes in Grade 2: 

```{r echo=TRUE}
reg_scores_g2 <- 
  main_data_g2 %>% filter(g2classtype=="REGULAR CLASS" | g2classtype=="REGULAR + AIDE CLASS", !is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc )

small_scores_g2 <- 
  main_data_g2 %>% filter(g2classtype=="SMALL CLASS", !is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc)

reg_scores_g2 <- 
  reg_scores_g2 %>% mutate(perc_maths_g2=percent_rank(g2tmathss), perc_read_g2 =percent_rank(g2treadss), perc_word_g2=percent_rank(g2wordskillss), perc_av_g2=percent_rank(g2aversc))

scores_g2 <- reg_scores_g2

for (i in 1:nrow(small_scores_g2)) {
  # select ith observation
  obs <- small_scores_g2[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g2, obs) %>% 
    mutate(perc_maths_g2=percent_rank(g2tmathss), perc_read_g2 =percent_rank(g2treadss), perc_word_g2=percent_rank(g2wordskillss),
perc_av_g2=percent_rank(g2aversc))
  # add observation to merged scores
  scores_g2 <-  add_row(scores_g2, tail(temp,1))
}


scores_g2 %>% 
  filter(g2classtype=="REGULAR CLASS" | g2classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_g2), mean_math = mean(perc_maths_g2), mean_read = mean(perc_read_g2), mean_word = mean(perc_word_g2))

```

Mean Scores for students in small classes in Grade 2:

```{r echo=TRUE}
scores_g2 %>% 
  filter(g2classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_g2, na.rm = TRUE), mean_math = mean(perc_maths_g2), mean_read = mean(perc_read_g2), mean_word = mean(perc_word_g2))
```

Mean Scores for students in regular classes in Grade 3:


```{r echo=TRUE}
reg_scores_g3 <- 
  main_data_g3 %>% filter(g3classtype=="REGULAR CLASS" | g3classtype=="REGULAR + AIDE CLASS", !is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid,g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid,g3aversc  )

small_scores_g3 <- 
  main_data_g3 %>% filter(g3classtype=="SMALL CLASS", !is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid, g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid,g3aversc  )

reg_scores_g3 <- 
  reg_scores_g3 %>% mutate(perc_maths_g3=percent_rank(g3tmathss), perc_read_g3 =percent_rank(g3treadss), perc_word_g3=percent_rank(g3wordskillss),  perc_av_g3=percent_rank(g3aversc))

scores_g3 <- reg_scores_g3

for (i in 1:nrow(small_scores_g3)) {
  # select ith observation
  obs <- small_scores_g3[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g3, obs) %>% 
    mutate(perc_maths_g3=percent_rank(g3tmathss), perc_read_g3 =percent_rank(g3treadss), perc_word_g3=percent_rank(g3wordskillss),  perc_av_g3=percent_rank(g3aversc))
  # add observation to merged scores
  scores_g3 <-  add_row(scores_g3, tail(temp,1))
}

scores_g3 %>% 
  filter(g3classtype=="REGULAR CLASS" | g3classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_av_g3), mean_math = mean(perc_maths_g3), mean_read = mean(perc_read_g3), mean_word = mean(perc_word_g3))

```
Mean Scores for students in small classes in Grade 3:

```{r echo=TRUE}
scores_g3 %>% 
  filter(g3classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_av_g3), mean_math = mean(perc_maths_g3), mean_read = mean(perc_read_g3), mean_word = mean(perc_word_g3))    
```

Finally, all percentile ranks are merged by student ID to get a dataset, which contails the ranks of every student in each grade:

```{r echo=TRUE}
# join all scores
scores_all <- 
  scores_gk %>% 
    full_join(scores_g1, by = "stdntid") %>% 
      full_join(scores_g2, by = "stdntid") %>% 
        full_join(scores_g3, by = "stdntid") %>% 
          select(stdntid, perc_av_g3, perc_maths_g3, perc_read_g3, perc_word_g3, perc_av_g2, perc_maths_g2, perc_read_g2, perc_word_g2, perc_av_g1, perc_maths_g1, perc_read_g1, perc_word_g1,perc_av_gk, perc_maths_gk, perc_read_gk, perc_word_gk)
# merge them with the main dataset
main_data <- 
  main_data %>% full_join(scores_all, by="stdntid")
```



#### Population percentile ranks

This approach computes the variables `perc2_[subject]_[grade]` directly from the whole population. 

Mean Scores for students in small classes in Kindergarden:

```{r echo=TRUE}
# Kindergarden
scores2_gk <- 
  main_data_gk %>% 
    filter(!is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) )  %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc) %>%   
      mutate(perc_fp_maths_gk=percent_rank(gktmathss), perc_fp_read_gk =percent_rank(gktreadss), perc_fp_word_gk=percent_rank(gkwordskillss), perc_fp_av_gk=percent_rank(gkaversc))

scores2_gk %>% 
  filter(gkclasstype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_gk), mean_math = mean(perc_fp_maths_gk), mean_read = mean(perc_fp_read_gk), mean_word = mean(perc_fp_word_gk))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_gk %>% 
  filter(gkclasstype=="REGULAR CLASS" | gkclasstype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_gk), mean_math = mean(perc_fp_maths_gk), mean_read = mean(perc_fp_read_gk), mean_word = mean(perc_fp_word_gk))
```  

Mean Scores for students in small classes in Grade 1:

```{r echo=TRUE}
# Kindergarden
scores2_g1 <- 
  main_data_g1 %>% 
    filter(!is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc) %>%   
      mutate(perc_fp_maths_g1=percent_rank(g1tmathss), perc_fp_read_g1 =percent_rank(g1treadss), perc_fp_word_g1=percent_rank(g1wordskillss), perc_fp_av_g1=percent_rank(g1aversc))

scores2_g1 %>% 
  filter(g1classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g1), mean_math = mean(perc_fp_maths_g1), mean_read = mean(perc_fp_read_g1), mean_word = mean(perc_fp_word_g1))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_g1 %>% 
  filter(g1classtype=="REGULAR CLASS" | g1classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g1), mean_math = mean(perc_fp_maths_g1), mean_read = mean(perc_fp_read_g1), mean_word = mean(perc_fp_word_g1))
``` 

Mean Scores for students in small classes in Grade 2:

```{r echo=TRUE}
# Kindergarden
scores2_g2 <- 
  main_data_g2 %>% 
    filter(!is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc) %>%   
      mutate(perc_fp_maths_g2=percent_rank(g2tmathss), perc_fp_read_g2 =percent_rank(g2treadss), perc_fp_word_g2=percent_rank(g2wordskillss), perc_fp_av_g2=percent_rank(g2aversc))

scores2_g2 %>% 
  filter(g2classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g2), mean_math = mean(perc_fp_maths_g2), mean_read = mean(perc_fp_read_g2), mean_word = mean(perc_fp_word_g2))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_g2 %>% 
  filter(g2classtype=="REGULAR CLASS" | g2classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g2), mean_math = mean(perc_fp_maths_g2), mean_read = mean(perc_fp_read_g2), mean_word = mean(perc_fp_word_g2))
```  

Mean Scores for students in small classes in grade 3:

```{r echo=TRUE}
# Kindergarden
scores2_g3 <- 
  main_data_g3 %>% 
    filter(!is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid,g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid, g3aversc) %>%   
      mutate(perc_fp_maths_g3=percent_rank(g3tmathss), perc_fp_read_g3 =percent_rank(g3treadss), perc_fp_word_g3=percent_rank(g3wordskillss), perc_fp_av_g3=percent_rank(g3aversc))

scores2_g3 %>% 
  filter(g3classtype=="SMALL CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g3), mean_math = mean(perc_fp_maths_g3), mean_read = mean(perc_fp_read_g3), mean_word = mean(perc_fp_word_g3))
```

And for those in regular classes:

```{r echo=TRUE}
scores2_g3 %>% 
  filter(g3classtype=="REGULAR CLASS" | g3classtype=="REGULAR + AIDE CLASS") %>%
    summarise(mean_av = mean(perc_fp_av_g3), mean_math = mean(perc_fp_maths_g3), mean_read = mean(perc_fp_read_g3), mean_word = mean(perc_fp_word_g3))
```  

Again, all percentile ranks are merged by student ID to get a dataset, which contains the ranks of every student in each grade:

```{r echo=TRUE}
# join all scores
scores_all2 <- 
  scores2_gk %>% 
    full_join(scores2_g1, by = "stdntid") %>% 
      full_join(scores2_g2, by = "stdntid") %>% 
        full_join(scores2_g3, by = "stdntid") %>% 
          select(stdntid, perc_fp_av_g3, perc_fp_maths_g3, perc_fp_read_g3, perc_fp_word_g3, perc_fp_av_g2, perc_fp_maths_g2, perc_fp_read_g2, perc_fp_word_g2, perc_fp_av_g1, perc_fp_maths_g1, perc_fp_read_g1, perc_fp_word_g1,perc_fp_av_gk, perc_fp_maths_gk, perc_fp_read_gk, perc_fp_word_gk)
# merge them with the main dataset
main_data <- 
  main_data %>% full_join(scores_all2, by="stdntid")
```


#### Percentile Ranks per School

This approach diverges from Krüger et al. 2001, but might be helpful since it directly controls variation on the school level. 

We expect percentile ranks for regular students within schools to follow a uniform distribution. Again the Variation in distribution of test scores from students in small classes is of interest.

```{r echo=TRUE}

small_scores_gk_sc <- 
  main_data_gk %>% filter(gkclasstype=="SMALL CLASS", !is.na(gktmathss) & !is.na(gktreadss) & !is.na(gkwordskillss) )  %>%
    select(stdntid,gkclasstype,gktmathss,gktreadss, gkwordskillss, gkschid, gkaversc  )


# We simply repeat the procedure with data grouped by school 
reg_scores_gk_sc <- 
  reg_scores_gk %>%
    group_by(gkschid) %>% 
      mutate(perc_maths_gk_sc=percent_rank(gktmathss), perc_read_gk_sc =percent_rank(gktreadss), perc_word_gk_sc=percent_rank(gkwordskillss), perc_av_gk_sc=percent_rank(gkaversc)) %>%
        ungroup()

scores_gk_sc <- reg_scores_gk_sc

for (i in 1:nrow(small_scores_gk_sc)) {
  # select ith observation
  obs <- small_scores_gk_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_gk_sc, obs) %>%
    group_by(gkschid) %>%
    mutate(perc_maths_gk_sc=percent_rank(gktmathss), perc_read_gk_sc =percent_rank(gktreadss), perc_word_gk_sc=percent_rank(gkwordskillss), perc_av_gk_sc=percent_rank(gkaversc)) %>% ungroup()
  # add observation to merged scores
  scores_gk_sc <-  add_row(scores_gk_sc, tail(temp,1))
}

# Grade 1
small_scores_g1_sc <- 
  main_data_g1 %>% filter(g1classtype=="SMALL CLASS", !is.na(g1tmathss) & !is.na(g1treadss) & !is.na(g1wordskillss) )  %>%
    select(stdntid,g1classtype,g1tmathss,g1treadss, g1wordskillss, g1schid, g1aversc)

reg_scores_g1_sc <- 
  reg_scores_g1 %>%
    group_by(g1schid) %>% 
      mutate(perc_maths_g1_sc=percent_rank(g1tmathss), perc_read_g1_sc =percent_rank(g1treadss), perc_word_g1_sc=percent_rank(g1wordskillss), perc_av_g1_sc=percent_rank(g1aversc))  %>%
        ungroup()

scores_g1_sc <- reg_scores_g1_sc

for (i in 1:nrow(small_scores_g1_sc)) {
  # select ith observation
  obs <- small_scores_g1_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g1_sc, obs) %>%
    group_by(g1schid) %>%
    mutate(perc_maths_g1_sc=percent_rank(g1tmathss), perc_read_g1_sc =percent_rank(g1treadss), perc_word_g1_sc=percent_rank(g1wordskillss), perc_av_g1_sc=percent_rank(g1aversc))  %>%
        ungroup()
  # add observation to merged scores
  scores_g1_sc <-  add_row(scores_g1_sc, tail(temp,1))
}

# Grade 2
small_scores_g2_sc <- 
  main_data_g2 %>% filter(g2classtype=="SMALL CLASS", !is.na(g2tmathss) & !is.na(g2treadss) & !is.na(g2wordskillss) )  %>%
    select(stdntid,g2classtype,g2tmathss,g2treadss, g2wordskillss, g2schid, g2aversc  )

reg_scores_g2_sc <- 
  reg_scores_g2 %>%
    group_by(g2schid) %>% 
      mutate(perc_maths_g2_sc=percent_rank(g2tmathss), perc_read_g2_sc =percent_rank(g2treadss), perc_word_g2_sc=percent_rank(g2wordskillss), perc_av_g2_sc=percent_rank(g2aversc))  %>%
        ungroup()

scores_g2_sc <- reg_scores_g2_sc

for (i in 1:nrow(small_scores_g2_sc)) {
  # select ith observation
  obs <- small_scores_g2_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g2_sc, obs) %>%
    group_by(g2schid) %>%
    mutate(perc_maths_g2_sc=percent_rank(g2tmathss), perc_read_g2_sc =percent_rank(g2treadss), perc_word_g2_sc=percent_rank(g2wordskillss), perc_av_g2_sc=percent_rank(g2aversc))  %>%
        ungroup()
  # add observation to merged scores
  scores_g2_sc <-  add_row(scores_g2_sc, tail(temp,1))
}


# Grade 3
small_scores_g3_sc <- 
  main_data_g3 %>% filter(g3classtype=="SMALL CLASS", !is.na(g3tmathss) & !is.na(g3treadss) & !is.na(g3wordskillss) )  %>%
    select(stdntid,g3classtype,g3tmathss,g3treadss, g3wordskillss, g3schid , g3aversc )

reg_scores_g3_sc <- 
  reg_scores_g3 %>%
    group_by(g3schid) %>% 
      mutate(perc_maths_g3_sc=percent_rank(g3tmathss), perc_read_g3_sc =percent_rank(g3treadss), perc_word_g3_sc=percent_rank(g3wordskillss), perc_av_g3_sc=percent_rank(g3aversc))  %>%
        ungroup()

scores_g3_sc <- reg_scores_g3_sc

for (i in 1:nrow(small_scores_g3_sc)) {
  # select ith observation
  obs <- small_scores_g3_sc[i,]
  # add to regular observations and compute percentile ranks
  temp <- add_row(reg_scores_g3_sc, obs) %>%
    group_by(g3schid) %>%
    mutate(perc_maths_g3_sc=percent_rank(g3tmathss), perc_read_g3_sc =percent_rank(g3treadss), perc_word_g3_sc=percent_rank(g3wordskillss), perc_av_g3_sc=percent_rank(g3aversc))  %>%
        ungroup()
  # add observation to merged scores
  scores_g3_sc <-  add_row(scores_g3_sc, tail(temp,1))
}


scores_all_sc <- 
  scores_gk_sc %>% 
    full_join(scores_g1_sc, by = "stdntid") %>% 
      full_join(scores_g2_sc, by = "stdntid") %>% 
        full_join(scores_g3_sc, by = "stdntid") %>% 
          select(stdntid, perc_av_g3_sc, perc_maths_g3_sc, perc_read_g3_sc, perc_word_g3_sc, perc_av_g2_sc, perc_maths_g2_sc, perc_read_g2_sc, perc_word_g2_sc, perc_av_g1_sc, perc_maths_g1_sc, perc_read_g1_sc, perc_word_g1_sc,perc_av_gk_sc, perc_maths_gk_sc, perc_read_gk_sc, perc_word_gk_sc)
# we get a problem here, because covariates are in different categories, better make a dataset with new variables and merge them with the main_data
main_data <- 
  main_data %>% full_join(scores_all_sc, by="stdntid")
```





# Repetition of previous analyzes



# Conceptual framework: FAMM

# Density analysis

# Implementation with refund

# To Do's and questions