---
title: "Analysis class performance as a collection of densities"
author: "Johannes Wagner"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
  toc: yes
code_folding: hide
---
  
  ```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=F, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```

```{r echo=FALSE}
library(tidyverse)
library(kdensity) # density estimation
library(lattice) # for histograms with relative frequencies
library(FDboost) # for functional analysis
```



```{r echo=TRUE}
load("/home/johannes/Documents/Projects/2020_Master_STAR/master20/data/main_data.RData")    
```


Model for School ID 205490 (with only 3 classes) and 63 observations.

```{r echo=TRUE}
main_data <- main_data %>% filter(gkschid==205490)

```


# Kernel Estimation

We use the apporach with beta-biased-kernels and user supplied bandwidth of `0.05` on overall percentiles for all students.

```{r echo=TRUE}
# split the data by class and then calculate a density for each class

# we want to filter the dataset for each class
# we sort the List by Teacher ID to get vectors in the right order
classList <- sort(unique(main_data$gktchid))
# make a list to store all densities
densities <- vector("list", length = length(classList)-1)

for (i in 1:length(classList)){
  temp <- main_data %>% 
    select(gktchid, perc_av_gk) %>% 
    filter(gktchid==classList[i])
  densities[[i]] <- kdensity(temp$perc_av_gk, start = "gumbel", kernel = "beta_biased", na.rm = T, bw = 0.08)}
```

Add covariates:
  
```{r echo=TRUE}
# create a list of classtypes as factor with 1=SMALL, 2=AID
listObjClasstype <- main_data %>% 
  # automatically orders tibble by gktchid to correspond to order in classList
  group_by(gktchid, gkclasstype) %>% 
  summarise(.groups = "drop") %>% 
  select(gkclasstype) %>%
  filter(!is.na(gkclasstype)) %>%
  unlist(use.names = F) %>%
  factor(labels=c("SMALL", "Regular", "wAid"))

# create a list of school IDs
listObjSchool <- main_data %>% 
  # automatically orders tibble by gktchid to correspond to order in classList
  group_by(gktchid, gkschid) %>% 
  summarise(.groups = "drop") %>% 
  select(gkschid) %>%
  filter(!is.na(gkschid)) %>%
  unlist(use.names = F) 

# comp_data_gender <- 
#   comp_data_gk %>% group_by(gktchid, gender) %>% 
#     summarise(n = n(), .groups="drop_last" ) %>% 
#       mutate(share_gender = n / sum(n)) %>% 
#         pivot_wider(id_cols =gktchid, names_from=gender, values_from= share_gender)
```

Transform Densities into **Design Matrix** 
  
```{r echo=TRUE}
numPoints = 100
# make a matrix where each row is a density and each column is an evaluation point
denMatrix <- matrix(data = NA, nrow = length(densities), ncol = numPoints)
eqdist <- seq(0,1,length.out = numPoints)  

for (i in 1:length(densities)){
  denMatrix[i,] <- densities[[i]](eqdist)
}  


# create a list with all variables for functional analysis  
star <- list(classDen_gk = denMatrix, 
             evPoints = eqdist,
             ClassType = listObjClasstype,
             School = listObjSchool) 



```


## Plot all class-specific densities

```{r echo=TRUE}
# par(mfrow = c(41,8), mar = c(2,2,2,2))
# for (i in 1:length(densities)){
#   plot(densities[[i]])
# }

## In one plot?
plot(densities[[1]], ylim = c(0,4.5))
for (i in 2:length(densities)){
  lines(densities[[i]])
} 


```

## Model with Classtype 

```{r echo=TRUE}
# Model with factor covariate region
mod_sc <- FDboost(classDen_gk ~ 1 + bolsc(ClassType, df = 4),
               timeformula = ~ bbs(evPoints),
               offset = "scalar", data = star)

plot(mod_sc)
# That is our model with regression lines for each school AND each Classtype
```


This result is very different from our model, where we included all schools, but fixed effects??

How is the effect on average?

```{r echo=TRUE}
  
main_data %>% group_by(gkclasstype) %>%
  summarize(mean = mean(perc_av_gk, na.rm = T))

```

Small class does much better in this school (due to lower **density** of bad students)