---
title: "Analysis class performance as a collection of densities"
author: "Johannes Wagner"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
  toc: yes
code_folding: hide
---
  
  ```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=F, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```

```{r echo=FALSE}
library(tidyverse)
library(kdensity) # density estimation
library(lattice) # for histograms with relative frequencies
library(FDboost) # for functional analysis
```



```{r echo=TRUE}
load("/home/johannes/Documents/Projects/2020_Master_STAR/master20/data/main_data.RData")    
```


# Kernel Estimation

We use the approach with beta-biased-kernels and user supplied bandwidth of `0.05` on overall percentiles for all students.

```{r echo=TRUE}
# split the data by class and then calculate a density for each class

# we want to filter the dataset for each class
# we sort the List by Teacher ID to get vectors in the right order
classList <- sort(unique(main_data$gktchid))
# make a list to store all densities
densities <- vector("list", length = length(classList))
# ignore first value of List since it is NA
for (i in 1:length(classList)){
  temp <- main_data %>% 
    select(gktchid, perc_av_gk) %>% 
    filter(gktchid==classList[i])
  densities[[i]] <- kdensity(temp$perc_av_gk, start = "gumbel", kernel = "beta_biased", na.rm = T, bw = 0.08)}
```

School-specific densities

```{r echo=TRUE}
# split the data by class and then calculate a density for each class

# we want to filter the dataset for each class
classList <- sort(unique(main_data$gktchid))
# make a list to store all densities
densities_sc <- vector("list", length = length(classList))
# ignore first value of List since it is NA
for (i in 1:length(classList)){
  temp <- main_data %>% 
    select(gktchid, perc_av_gk_sc) %>% 
    filter(gktchid==classList[i])
    densities_sc[[i]] <- kdensity(temp$perc_av_gk, start = "gumbel", kernel = "beta_biased", na.rm = T, bw = 0.08)}
```

Add covariates:

```{r echo=TRUE}
# create a list of classtypes as factor with 1=SMALL, 2=AID
listObjClasstype <- main_data %>% 
  # automatically orders tibble by gktchid to correspond to order in classList
  group_by(gktchid, gkclasstype) %>% 
  summarise(.groups = "drop") %>% 
  select(gkclasstype) %>%
  filter(!is.na(gkclasstype)) %>%
  unlist(use.names = F) %>%
  factor(labels=c("SMALL", "Regular", "wAid"))
  
# create a list of school IDs
listObjSchool <- main_data %>% 
  # automatically orders tibble by gktchid to correspond to order in classList
  group_by(gktchid, gkschid) %>% 
  summarise(.groups = "drop") %>% 
  select(gkschid) %>%
  filter(!is.na(gkschid)) %>%
  unlist(use.names = F) 

# comp_data_gender <- 
#   comp_data_gk %>% group_by(gktchid, gender) %>% 
#     summarise(n = n(), .groups="drop_last" ) %>% 
#       mutate(share_gender = n / sum(n)) %>% 
#         pivot_wider(id_cols =gktchid, names_from=gender, values_from= share_gender)
```

Transform Densities into **Design Matrix** 

```{r echo=TRUE}
numPoints = 100
# make a matrix where each row is a density and each column is an evaluation point
denMatrix_1 <- matrix(data = NA, nrow = length(densities), ncol = numPoints)
# for school specific densities
denMatrix_2 <- matrix(data = NA, nrow = length(densities_sc), ncol = numPoints)

eqdist <- seq(0,1,length.out = numPoints)  

for (i in 1:length(densities)){
  denMatrix_1[i,] <- densities[[i]](eqdist)
  denMatrix_2[i,] <- densities_sc[[i]](eqdist)
}  


# create a list with all variables for functional analysis  
star <- list(classDen_gk = denMatrix_1, 
             classDen_gk_sc = denMatrix_2,
             evPoints = eqdist,
             ClassType = listObjClasstype,
             School = listObjSchool) 



```


## Plot all class-specific densities

```{r echo=TRUE}
# par(mfrow = c(41,8), mar = c(2,2,2,2))
# for (i in 1:length(densities)){
#   plot(densities[[i]])
# }

## In one plot?
plot(densities[[1]], ylim = c(0,4.5))
for (i in 2:length(densities)){
  lines(densities[[i]])
} 


```

It seems we have quite a lot of "extreme" classes with either a very skewed distribution to the left or to the right (more so to the left it seems)

## Reproducing Histogram Representation (Eva)

We show histogram for percentiles given Grade and Class Size:

```{r echo=TRUE}
smallClassesGK <- main_data %>% 
  select(perc_av_gk, gkclasstype) %>% 
  filter(gkclasstype=="SMALL CLASS" & !is.na(perc_av_gk)) 
regClassesGK <- main_data %>% 
  select(perc_av_gk, gkclasstype) %>% 
  filter(gkclasstype=="REGULAR CLASS" & !is.na(perc_av_gk)) 
regAidClassesGK <- main_data %>% 
  select(perc_av_gk, gkclasstype) %>% 
  filter(gkclasstype=="REGULAR + AIDE CLASS" & !is.na(perc_av_gk)) 
```

### Plot Histogramms for groups

```{r echo=TRUE}
par(mfrow = c(1,3))
h1 <- hist(smallClassesGK$perc_av_gk, plot=F, breaks = 20)
h1$counts <- h1$counts / sum(h1$counts)
plot(h1, freq=TRUE, ylab="Relative Frequency", ylim = c(0,0.2))
h2 <- hist(regClassesGK$perc_av_gk, plot = F, breaks = 20)
h2$counts <- h2$counts / sum(h2$counts)
plot(h2, freq=TRUE, ylab="Relative Frequency", ylim = c(0,0.2))
h3 <- hist(regAidClassesGK$perc_av_gk, plot = F, breaks = 20)
h3$counts <- h3$counts / sum(h3$counts)
plot(h3, freq=TRUE, ylab="Relative Frequency", ylim = c(0,0.2))

```

We can see that small Classes decrease the densities of percentiles betwen 0 and 0.8 (homogenous effect) and increase their desnities for percentiles higher than 0.8.

A small effect seems to be that Aids decrease the number of students in the lowest percentile and therefore slightly increasing the densities of percentiles between 0.1 and 0.5. **This might be a relevant effect, which needs clearification**.

### Show group specific densities (later: with different grades in color)

```{r echo=TRUE}
# we want to filter the dataset for each class
groupList <- unique(main_data$gkclasstype)
# make a list to store all densities
SizeTypedensities <- vector("list", length = length(groupList)-1)
# ignore first value of List since it is NA
for (i in 2:length(groupList)){
  temp <- main_data %>% 
    select(gkclasstype, perc_av_gk) %>% 
    filter(gkclasstype==groupList[i])
  SizeTypedensities[[i-1]] <- kdensity(temp$perc_av_gk, start = "gumbel", kernel = "beta_biased", na.rm = T, bw = 0.05)}  


par(mfrow = c(1,3))
plot(SizeTypedensities[[1]], ylim = c(0,1.4), main = "with Aid")
plot(SizeTypedensities[[2]], ylim = c(0,1.4), main = "Small Classes")
plot(SizeTypedensities[[3]], ylim = c(0,1.4), main = "Regular Classes")

```

As we can see, densities are are more pronounced in representing the effects mentioned before.

Results are strongly depending on the choice of bandwidth

## Reproducing Histogram using school-specific class densities



### Plot Histogramms for groups

We show histogram for percentiles given Grade and Class Size:

```{r echo=TRUE}
smallClassesGK <- main_data %>% 
  select(perc_av_gk_sc, gkclasstype) %>% 
  filter(gkclasstype=="SMALL CLASS" & !is.na(perc_av_gk_sc)) 
regClassesGK <- main_data %>% 
  select(perc_av_gk_sc, gkclasstype) %>% 
  filter(gkclasstype=="REGULAR CLASS" & !is.na(perc_av_gk_sc)) 
regAidClassesGK <- main_data %>% 
  select(perc_av_gk_sc, gkclasstype) %>% 
  filter(gkclasstype=="REGULAR + AIDE CLASS" & !is.na(perc_av_gk_sc)) 
```



```{r echo=TRUE}
par(mfrow = c(1,3))
h1 <- hist(smallClassesGK$perc_av_gk_sc, plot=F, breaks = 20)
h1$counts <- h1$counts / sum(h1$counts)
plot(h1, freq=TRUE, ylab="Relative Frequency", ylim = c(0,0.2))
h2 <- hist(regClassesGK$perc_av_gk_sc, plot = F, breaks = 20)
h2$counts <- h2$counts / sum(h2$counts)
plot(h2, freq=TRUE, ylab="Relative Frequency", ylim = c(0,0.2))
h3 <- hist(regAidClassesGK$perc_av_gk_sc, plot = F, breaks = 20)
h3$counts <- h3$counts / sum(h3$counts)
plot(h3, freq=TRUE, ylab="Relative Frequency", ylim = c(0,0.2))

```

After controlling for school effects, the small Class effect is stronger, while the Aid effect is weaker.

### Show group specific densities (later: with different grades in color)

```{r echo=TRUE}
# make a list to store all densities
SizeTypeDensities_sc <- vector("list", length = length(groupList)-1)
# ignore first value of List since it is NA
for (i in 2:length(groupList)){
  temp <- main_data %>% 
    select(gkclasstype, perc_av_gk_sc) %>% 
    filter(gkclasstype==groupList[i])
  SizeTypeDensities_sc[[i-1]] <- kdensity(temp$perc_av_gk_sc, start = "gumbel", kernel = "beta_biased", na.rm = T, bw = 0.3)}  
par(mfrow = c(1,3))
plot(SizeTypeDensities_sc[[1]], ylim = c(0,1.4), main = "with Aid")
plot(SizeTypeDensities_sc[[2]], ylim = c(0,1.4), main = "Small Classes")
plot(SizeTypeDensities_sc[[3]], ylim = c(0,1.4), main = "Regular Classes")

```

It seems strange that the data becomes less clear after we control for school-fixed effects (maybe problematic that we first compute percentiles on a school level and afterward looking for effects on a more general level?)

# Next Step: Repeat first Line of argument with groups for each School and color for four components (that i smost relevent for **school-specific-densities**)
.

# Next Step: Transform densities into clr coordinates

Do we use a simple L2 space or Bayes Hilbert space??

# Next Step: Find optimal bandwidth trough calculating "unbiased cross validation" for each group

use: https://bookdown.org/egarpor/NP-UC3M/kde-i-bwd.html


# Regression Model



### Intercept Model

```{r echo=TRUE}
# Intercept Model with standard defaults 
m0 <- FDboost(classDen_gk ~ 1,
              timeformula = ~ bbs(evPoints),
              offset = "scalar", data = star)
# Error in bsplines(mf[[i]], knots = args$knots[[i]]$knots, boundary.knots = args$knots[[i]]$boundary.knots,  : 
#  some ‘x’ values are beyond ‘boundary.knots’
# plot(m0)  
```


## Model1 with Covariates

Model 1: Only Classtype

```{r echo=TRUE}
# Model with factor covariate region
mod1 <- FDboost(classDen_gk ~ 1 + bolsc(ClassType, df = 4),
               timeformula = ~ bbs(evPoints),
               offset = "scalar", data = star)

# plot(mod1)
```

Model 1b: Classtype for school specific densities
```{r echo=TRUE}
# Model with factor covariate region
mod1b <- FDboost(classDen_gk_sc ~ 1 + bolsc(ClassType, df = 4),
               timeformula = ~ bbs(evPoints),
               offset = "scalar", data = star)

# plot(mod1b)
```

This validates our OLS results in the sense that only SMALL class matters. 


Model 2:

with Classtype and School Effects

```{r echo=TRUE}
# Model with factor covariate region
mod2 <- FDboost(classDen_gk ~ 1 +  bolsc(School, df = 4) + bolsc(ClassType, df = 4),
               timeformula = ~ bbs(evPoints),
               offset = "scalar", data = star)

# plot(mod2)
# That is our model with regression lines for each school AND each Classtype

```

**COMMENT**: Why is there no difference in the coefficients for Classtype when we include random effects for schools?? 
- actually there is: the effects are much smaller in size (around |0.3| vs |0.15|) -> cp. to regression model where coefficients dont change -> here relative influence doesn't change
- OLS Model gives coefficient between 0.05 and 0.1 on average


```{r echo=TRUE}
# Model with factor covariate region
mod2b <- FDboost(classDen_gk_sc ~ 1 +  bolsc(School, df = 4) + bolsc(ClassType, df = 4),
               timeformula = ~ bbs(evPoints),
               offset = "scalar", data = star)

# plot(mod2b)
# That is our model with regression lines for each school AND each Classtype

```

Again, same results with smaller coefficients

## Model for selected schools

First, can we identify the school with biggest share of "bad" students?

```{r echo=TRUE}
test <- main_data %>% group_by(gkschid) %>%
  summarise(mean_perc = mean(perc_av_gk, na.rm = T)) %>%
  arrange(mean_perc)
# School ID 205490 is a massive underperformer
```

Model for School ID 205490




Actually interesting that we see one pressumbly **Outlier** school here (blue). Why dont we see coefficient for Classtype?

## Comments

Theoretically densities should represent the relative frequencies of observations in the related group, i.e. Small Class in school x.


First results imply that the improvement of students skils on average is due to a strong effect in the **top quintile** of the class percentile distribution (within schools). I.e. strong students get better on average than bad students that are less bad. 
