---
title: "Density-on-scalar regression for class performence"
author: "Johannes Wagner"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: yes
    code_folding: hide
---
  
```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=F, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```

```{r echo=FALSE}
library(tidyverse)
library(kdensity) # density estimation
library(lattice) # for histograms with relative frequencies
library(FDboost) # for functional analysis
library(compositions) # for clr transformation
```

# Density-on-scalar estimation

## Prerequisites 

## Density estimation


Following a first density-on-scalar estimation for the STAR data is represented.
So far the model only accounts for (well known) effects of assigned classtypes amd only analyses students in kindergarten.

We use data that consists of a list of
- the densities (kernel estimation from chapter 3) for the percentiles for average score results for students in kindergarten
- a vector of evaluation points that were used to retrieve observation points for the densities
- a vector of classtypes with levels: "SMALL", "Reg", "wAid"
- a dummy vector indicating the assignment of a SMALL class
- a vector of school specific indicators with 79 levels

```{r echo=TRUE}
load("/home/johannes/Documents/Projects/2020_Master_STAR/master20/data/functionalData.RData")    
```

## Representation of clr-transformed densities

At first, we can plot all class-specific densities into one plot and for specific groups. There are a few extreme classes with extremly skewed distrinutions to their left tails. To make the plots more relatable we restrict the y-axis for the group specific plots on values between -50 and 50 log ratios.


```{r echo=TRUE}
# get the densities
depDens <- star$classDen_gk
evPoints <- star$evPoints
plot(evPoints,as.double(depDens[1,]), type="l", ylim = c(-800,100))
for (i in 2:nrow(depDens)){
  lines(evPoints,as.double(depDens[i,]), type="l", ylim = c(-800,100))
}   
```

There are a few "extreme" classes with an extreme underrepresentation of students with very low percentiles. We can avoid this by calculating school-specific densities.

```{r echo=TRUE}
# get the densities
depDens <- star$classDen_gk_sc
evPoints <- star$evPoints
classType <- star$ClassType
plot(evPoints,as.double(depDens[1,]), type="l", ylim = c(-200,100))
for (i in 2:nrow(depDens)){
  lines(evPoints,as.double(depDens[i,]), type="l", ylim = c(-200,100))
}   
```

Nevertheless, it appears that there are four extreme classes, which we will analyse later.
What we will do at this point is to present the densities for specific groups.

```{r echo=TRUE}
subDepDens <- depDens[classType=="SMALL",]
plot(evPoints,as.double(subDepDens[1,]), type="l", ylim = c(-50,50), main="Small classes")
for (i in 2:nrow(subDepDens)){
  lines(evPoints,as.double(subDepDens[i,]), type="l", ylim = c(-50,50))
}    

```
It appears that all extreme classes are classes of the subset of small classes and therefor we can expect that this influences a mean regression procedure quite heavily.

```{r echo=TRUE}
subDepDens <- depDens[classType=="Regular",]
plot(evPoints,as.double(subDepDens[1,]), type="l", ylim = c(-50,50), main="Reg classes")
for (i in 2:nrow(subDepDens)){
  lines(evPoints,as.double(subDepDens[i,]), type="l", ylim = c(-50,50))
}    

```

The amount of variation for regular classes is much smaller than for SMALL classes, but again this is probably heavily influenced by the extreme classes.

```{r echo=TRUE}
subDepDens <- depDens[classType=="wAid",]
plot(evPoints,as.double(subDepDens[1,]), type="l", ylim = c(-50,50), main="with Aides")
for (i in 2:nrow(subDepDens)){
  lines(evPoints,as.double(subDepDens[i,]), type="l", ylim = c(-50,50))
}    
```

Those Plots are probably not very helpful though. **IGNORE**

## Model Specification 

Starting with the regression of (school-specific) class densities onto the classtype while controlling for schools, we estimate several possible Specifications for the regression problem going from simple to more complex models.  

$$
\begin{array}{l}
\operatorname{clr}\left[\hat{f}_{\text {school, classtype, grade=kindergarten }}\right]=\operatorname{clr}\left[\beta_{0}\right]+\operatorname{clr}\left[\beta_{\text {SMALL }}\right] \cdot \mathbb{1}_{\text {reg_small=small}}(\text { classtype })+\operatorname{clr}\left[\beta_{\text {classtype }}\right] \\
\quad+\operatorname{clr}\left[\beta_{\text {school }}\right]+\operatorname{clr}[g(\text { school})]+\operatorname{clr}\left[\varepsilon_{\text {region, child group, year }}\right]
\end{array}
$$
