---
title: "EDA"
author: "Johannes Wagner"
date: "9/12/2020"
output: 
  html_document:
    toc: yes
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, include=FALSE}
library(tidyverse)
```

# Overview Dataset STAR

We load the data from [harvard.com](https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/10766)

```{r echo=FALSE, include=FALSE}
load("/home/hanes/Documents/Projects/2020_Master_STAR/master20/data/STAR_Students.RData")

```

- 11.602 Students by ID

## Core Data

- Demographics
- School and class identifiers
- school teacher information
- Experimental condition (“class type”)
- Norm-referenced and criterion-referenced achievement test scores
- Motivation and self-concept scores.

### Additional Data

- Achievement test scores for the students when they were in grades 4 – 8, obtained from the Tennessee State Department of Education
- Teachers’ ratings of student behavior in grades 4 and 8
- Students’ self-reports of school engagement and peer effects in grade 8
- Course taking in mathematics, science, and foreign language in high school, obtained from student transcripts
- SAT/ACT participation and scores, obtained from ACT, Inc. and from Educational Testing Service
- Graduation/dropout information, obtained from high school transcripts and the Tennessee State Department of Education



## First check

Variables by name and values

All variables:
```{r echo=TRUE}
list <- attributes(x)
list$orig.names

```


## Core Variables: Demographics
### ID
```{r echo=TRUE}
x %>% summarize(min = min(stdntid, na.rm = TRUE),max = max(stdntid, na.rm = TRUE), num_NA = sum(is.na(stdntid)) )   
```
No missings.

### Gender
```{r echo=TRUE}
summary(x$gender)
attributes(x$gender)
```

### race

```{r echo=TRUE}
summary(x$race)
attributes(x$race)
```

### free lunch

```{r echo=TRUE}
# Kindergarten
summary(x$gkfreelunch)
attributes(x$gkfreelunch)
# Grade 1
summary(x$g1freelunch)
attributes(x$g1freelunch)
# Grade 2
summary(x$g2freelunch)
attributes(x$g2freelunch)
# Grade 3
summary(x$g3freelunch)
attributes(x$g3freelunch)
```

## Performance

### Repetition of class

```{r echo=TRUE}
# Kindergarten
summary(x$gkrepeat)
attributes(x$gkrepeat)
```

### Special Education
```{r echo=TRUE}
# Kindergarten
summary(x$gkspeced)
attributes(x$gkspeced)
# Grade 1
summary(x$g1speced)
attributes(x$g1speced)
# Grade 2
summary(x$g2speced)
attributes(x$g2speced)
# Grade 3
summary(x$g3speced)
attributes(x$g3speced)
```

### Reading Skills

```{r echo=TRUE}
# Kindergarten
summary(x$gktreadss)
attributes(x$gktreadss)
# Grade 1
summary(x$g1treadss)
attributes(x$g1treadss)
# Grade 2
summary(x$g2treadss)
attributes(x$g2treadss)
# Grade 3
summary(x$g3treadss)
attributes(x$g3treadss)

```

### Self evalaluation

```{r echo=TRUE}
# Kindergarten
summary(x$gkmotivraw)
attributes(x$gkmotivraw)
# Grade 1
summary(x$g1motivraw)
attributes(x$g1motivraw)
# Grade 2
summary(x$g2motivraw)
attributes(x$g2motivraw)
# Grade 3
summary(x$g3motivraw)
attributes(x$g3motivraw)
```

### Math Skills

```{r echo=TRUE}
# Kindergarten
summary(x$gktmathss)
attributes(x$gktmathss)
# Grade 1
summary(x$g1tmathss)
attributes(x$g1tmathss)
# Grade 2
summary(x$g2tmathss)
attributes(x$g2tmathss)
# Grade 3
summary(x$g3tmathss)
attributes(x$g3tmathss)

```

### Listening Skills

```{r echo=TRUE}
# Kindergarten
summary(x$gktlistss)
attributes(x$gktlistss)
# Grade 1
summary(x$g1tlistss)
attributes(x$g1tlistss)
# Grade 2
summary(x$g2tlistss)
attributes(x$g2tlistss)
# Grade 3
summary(x$g3tlistss)
attributes(x$g3tlistss)

```

### Word Skills

```{r echo=TRUE}
# Kindergarten
summary(x$gkwordskillss)
attributes(x$gkwordskillss)
# Grade 1
summary(x$gkwordskillss)
attributes(x$gkwordskillss)
# Grade 2
summary(x$gkwordskillss)
attributes(x$gkwordskillss)
# Grade 3
summary(x$gkwordskillss)
attributes(x$gkwordskillss)
```

### Motivation

```{r echo=TRUE}
# Kindergarten
summary(x$gkmotivraw)
attributes(x$gkmotivraw)
# Grade 1
summary(x$gkmotivraw)
attributes(x$gkmotivraw)
# Grade 2
summary(x$gkmotivraw)
attributes(x$gkmotivraw)
# Grade 3
summary(x$gkmotivraw)
attributes(x$gkmotivraw)
```

### SAT/ACT Scores

SAT 
```{r echo=TRUE}
# Math
summary(x$hssatmath)
attributes(x$hssatmath)
# total
summary(x$hssattot)
attributes(x$hssattot)
```

ACT 
```{r echo=TRUE}
# Math
summary(x$hsactmath)
attributes(x$hsactmath)
# total
summary(x$hssattot)
attributes(x$hssattot)
```


## Kindergarten characteristics 

### Absent time
```{r echo=TRUE}
summary(x$gkabsent)
attributes(x$gkabsent)
```

### urban/rural
```{r echo=TRUE}
summary(x$gksurban)
attributes(x$gksurban)
```

### years in kindergarten
```{r echo=TRUE}
summary(x$gktyears)
attributes(x$gktyears)
```

## School characteristics

### school ID
```{r echo=TRUE}
summary(x$g1schid)
attributes(x$g1schid)
```

### urban/rural
```{r echo=TRUE}
summary(x$g1schid)
attributes(x$g1schid)
```

## Teacher characteristics

### Gender
```{r echo=TRUE}
summary(x$g1tgen)
attributes(x$g1tgen)
```

### High School Degree
```{r echo=TRUE}
summary(x$g1thighdegree)
attributes(x$g1thighdegree)
```

### Race
```{r echo=TRUE}
summary(x$g1trace)
attributes(x$g1trace)
```

### Career Level
```{r echo=TRUE}
summary(x$g1tcareer)
attributes(x$g1tcareer)
```

### Experience Level
```{r echo=TRUE}
summary(x$g1tyears)
attributes(x$g1tyears)
```





## Core Variables: Design 

### Level of Entry/Participation
```{r echo=TRUE}
# Kindergarten
summary(x$X)
attributes(x$X)
# Grade 1
summary(x$X.1)
attributes(x$X.1)
# Grade 2
summary(x$X.2)
attributes(x$X.2)
# Grade 3
summary(x$X.3)
attributes(x$X.3)
```

### Flags for achievement Data 
```{r echo=TRUE}
# Kindergarten
summary(x$flaggk)
# Grade 1
summary(x$flagg1)
# Grade 2
summary(x$flagg2)
# Grade 3
summary(x$flagg3)
# Grade 4
summary(x$flagg4)
# Grade 5
summary(x$flagg5)
# Grade 6
summary(x$flagg6)
# Grade 7
summary(x$flagg7)
# Grade 8
summary(x$flagg8)
```
### Flags for further participation Studies
```{r echo=TRUE}
# Grade 4
summary(x$flagprt4)    
# Grade 8: self identification
summary(x$flagidn8)
# Grade 8: participation
summary(x$flagprt8)
# University entry: SAT/ACT
summary(x$flagsatact)
# Highschool Data 
summary(x$flaghscourse)
# Highschool Data Graduation
summary(x$flaghsgraduate)
```
### Class Room Type (**core Treatment**)

```{r echo=TRUE}
# Kindergarten
summary(x$gkclasstype)
attributes(x$gkclasstype)
summary(x$gkclasssize)
attributes(x$gkclasssize)
# Grade 1
summary(x$g1classtype)
attributes(x$g1classtype)
summary(x$g1classsize)
attributes(x$g1classsize)
# Grade 2
summary(x$g2classtype)
attributes(x$g2classtype)
summary(x$g2classsize)
attributes(x$g2classsize)
# Grade 3
summary(x$g3classtype)
attributes(x$g3classtype)
summary(x$g3classsize)
attributes(x$g3classsize)
```
### Flags for Composition (mobility of students)

```{r echo=TRUE}
summary(x$cmpstype)
attributes(x$cmpstype)
summary(x$cmpsdura)
attributes(x$cmpsdura)
summary(x$yearsstar)
attributes(x$yearsstar)
summary(x$yearssmall)
attributes(x$yearssmall)
```



## Probably not relevant

### birthmonth, birthday
```{r echo=TRUE}
summary(x$birthmonth)
attributes(x$birthmonth)
summary(x$birthday)
attributes(x$birthday)
```
