---
title: "EDA"
author: "Johannes Wagner"
date: "9/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview Dataset STAR

We load the data from [harvard.com](https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/10766)

```{r echo=FALSE, include=FALSE}
load("/home/hanes/Documents/Projects/2020_Master_STAR/Data/STAR_Students.RData")
```

- 11.602 Students by ID

## Core Data

- Demographics
- School and class identifiers
- school teacher information
- Experimental condition (“class type”)
- Norm-referenced and criterion-referenced achievement test scores
- Motivation and self-concept scores.

### Additional Data

- Achievement test scores for the students when they were in grades 4 – 8, obtained from the Tennessee State Department of Education
- Teachers’ ratings of student behavior in grades 4 and 8
- Students’ self-reports of school engagement and peer effects in grade 8
- Course taking in mathematics, science, and foreign language in high school, obtained from student transcripts
- SAT/ACT participation and scores, obtained from ACT, Inc. and from Educational Testing Service
- Graduation/dropout information, obtained from high school transcripts and the Tennessee State Department of Education